Parallel Information Set Generation for Kriegspiel

### 2012-07-12 sbac/paper.tex

The paper is organized as follows.  In Section~\ref{Background}, we review
necessary background material, including the rules of Kriegspiel and the nature
of the information set generation problem.  In Section ~\ref{ParSSSE}, we
describe the {\sc Charm++} Parallel State Space Search Engine (ParSSSE)
platform.  We then provide our experimental results in Section~\ref{Results}.
Related work is discussed in Section~\ref{Related}.  We conclude with a
discussion of future work.

%\Section{Information Set Generation}
\Section{Kriegspiel}
Information Set Generation is the process of identifying the nodes in an
imperfect information game tree that are consistent with a player's
observations~\cite{richards12information}. Here, we present our techniques and
results in the context of a particular game: Kriegspiel~\cite{li94chess}.
However, the techniques and framework presented in this paper are general and
can be applied to ISG in other imperfect information games.

%\Section{Background: Kriegspiel}
Kriegspiel, or partially observable chess,  was invented by Henry Michael
Temple in 1899.  Historically, the game has required three parties: two
competing players and one referee.  Today, players often compete over a
computer network, with the role of the referee being automated.  One player
sees and controls only the white pieces; the other, black.  The referee tracks
both sets of pieces.  Instead of openly declaring their moves, players secretly
communicate move requests to the referee.  The referee, who knows the location
of all pieces, assesses the legality of the move according to the standard
rules of chess.

@@

\Section{Information Set Generation}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figure~\ref{listing} and show a full listing and the associated pictorial progression of an actual
Kriegspiel game~\cite{li94chess}.  We have adapted the notation used by Wolfe for Berkeley Kriegspiel.  Odd (resp. even) moves are for white (resp. black).  The actual move is shown
first followed by a list of referee announcements.  Captures announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn tries announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.  For example at move 9, (Figure~\ref{fig:9W}--\ref{fig:9B}),
black attempts to block a potential threat along the diagonal by moving a pawn to f6, then attempts to escape or capture
a threat at f7.  When both of these attempts fail, black can infer that his king is being threatened by a protected
bishop or queen at f7.  The attempt to move to e7 is black's way of finding out whether the threatening piece is a queen
or bishop.  When this move fails, black knows it is white's queen and ultimately retreats to d8.  Note that the failed
moves by white on turns 3 and 4 are a result of being blocked by the pawn at e2.

@@

\begin{figure}[htpb]
\small
%\begin{algorithm}
\begin{boxedminipage}{\columnwidth}
\begin{algorithmic}[1]
\Function{DFS-Infoset}{$a_{1:t}$, $o_{1:T}^{(k)}$, $I$, $s$, $t$}
  \If {$t > T$}
    \State {$I \leftarrow I \cup a_{1:t}$} %\Comment{$a_{1:t}$ leads to $o_{1:T}$ for $P_k$; add it to the information set.}
  \Else {
    \For{\textbf{each} legal action $a$ at $s$}
      \State {$s'$ = {\sc GetNextState}($s$,$a$)}\label{line:update}
      \If {$s'.o_{t+1}^{(k)}=o_{t+1}^{(k)}$}
        \State {{\sc DFS-Infoset}($a_{1:t}a$, $o_{1:T}^{(k)}$, $I$, $s'$, $t+1$)} 
          %\Comment{ Recall that $N(a_{1:t}a) = T(T(a_{1:t}),a)$}
      \EndIf
    \EndFor
  }\EndIf
\EndFunction
\label{codelisting}
\end{algorithmic}
\end{boxedminipage}
\caption{Algorithm}
\end{figure}

\Section{Parallel Implementation using Charm++ and ParSSSE}\label{ParSSSE}
We implemented our search engine on top of the {\sc Charm++} run-time
system for parallel processing~\cite{kale93charm,kale09charm}.  {\sc Charm++}
is an extension to the C++ programming language that provides
machine-independent infrastructure for parallel computation.  The language and
its accompanying runtime system have been ported to many shared-memory and
distributed-memory platforms.

@@

\Section{Performance Results}\label{Results}
In order to understand the effect of various parameters such as grainsize and
load balancing strategy, we first evaluated the scalability of our
implementation using a randomly selected problem instance. Next, we show that
our implementation performs equally well for instances from actual games.

\subsection{Analysis using Random Instance} 
The performance results reported in this section were obtained by running our
program on Blueprint with up to 1,024 processors. TODO: Blueprint H/w or
citation

@@

\begin{table}
\centering
\begin{tabular}{rrr}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}
\vspace{-0.1in}

@@

\label{speedups}
\vspace{-0.3in}
\end{table}

\subsubsection{Effect of Load Balancing}
\begin{figure}[h]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}
\vspace{-0.1in}
\caption{Effect of load-balancing strategy}
\label{3schemes}
\vspace{-0.2in}
\end{figure}

In our next round of experiments, we set out to compare load-balancing
strategies for the problem.  We used the random, work stealing, and
neighborhood balancing strategies that are available in the general purpose
search engine.  These results are shown in Figure~\ref{3schemes}.  Our results
show that there was actually little variation in performance across the
different strategies.  Across all of our experiments, the work stealing
approach appeared to be consistently---if only marginally---better than the
others.  Our hypothesis here is that the nodes are rarely starved for work.  In
other words, perhaps the very nature of the problem means that the load is
naturally well-balanced and that the load-balancer rarely needs to do a
significant amount of redistribution.  This is probably also related to the
fact that we are doing an all-solutions search.

\subsubsection{Effect of Problem Size}
\begin{figure}[h]
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depths.pdf}
\end{minipage}
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depthsnolog.pdf}
\end{minipage}
\vspace{-0.1in}

@@

\vspace{-0.2in}
\label{3depths}
\end{figure}

@@

\begin{figure*}[!ht]
\centering
\subfigure[Depth 8]{
\includegraphics[width=2.2in]{plots/dep8.pdf}
\label{dep8}
} 
\subfigure[Depth 10]{
\includegraphics[width=2.2in]{plots/dep10.pdf}
\label{dep10}
} 
\subfigure[Depth 12]{
\includegraphics[width=2.2in]{plots/dep12.pdf}
\label{dep12}
} 
\vspace{-0.1in}
\caption{Grainsize Control for different problem sizes}
%\vspace{-0.1in}
\label{fig:dep}
\vspace{-0.1in}
\end{figure*}

\subsection{Real Game Instance}
\begin{figure}[ht]
 \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem4.pdf} %

\caption{Kriegspiel position from Virgil-David 1969 with information set size
764,209.}

\label{prob4} 
\vspace{-0.2in}
\end{figure}

\begin{figure}[ht]
 \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem5.pdf} %

\caption{Kriegspiel position from Game 6 of Kbot-Darkboard at the 2006 Computer
Chess championship.  Information set size is 5,099,257.}

\label{prob5} 
\vspace{-0.2in}
\end{figure}

@@

\begin{table}[tphb]
\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 1844$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

\begin{table}[tphb]
\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & 617$|$1 & {\bf 612$|$1} & 650$|$0.95 & 3312$|$0.19 \\
4 & 213$|$2.89 & 170$|$3.63 & {\bf 158$|$3.89} & 367$|$1.68 \\
16 & 83$|$7.41 & 49$|$12.4 & {\bf 40$|$15.1} & 64$|$9.57 \\
64 & 54$|$11.3 & 19$|$32.3 & {\bf 11$|$52.9} & 14.9$|$41.4 \\
256 & 54$|$11.3 & 9.8$|$62.9 & 3.5$|$172 & {\bf 3.5$|$174} \\
1024 & 58$|$10.6 & 5.5$|$111 & 1.3$|$464 & {\bf 1.2$|$536} \\
\end{tabular}

@@

@@

\begin{table}[tphb]
\centering
\begin{tabular}{lrrc}
{\bf Authors} & {\bf Game} & {\bf Year} & {\bf Best Speedup (cores)} \\
Ferguson \& Korf & Othello & 1988 & 12(32) \\
Felton \& Otto & Chess & 1988 & 100(256) \\
Richards \etal & Kriegspiel & 2012 & 178(256) \\
Richards \etal & Kriegspiel & 2012 & {\bf 536(1024)} \\
\end{tabular}
\caption{Best Speedups reported for game tree applications.}
\vspace{-0.2in}
\label{bestspeedups}
\end{table}

\Section{Related Work}
An overview of the theory of imperfect information games and other basic topics
in game theory can be found in~\cite{kuhn03lectures} and~\cite{kuhn97classics}.
A technical explanation of the concept of information sets can be found
in~\cite{gilpin07algorithms}.  Koller \etal, showed that a Nash equilibrium can
be found in time that is polynomial in the number of nodes in the game
tree~\cite{koller94fast}.  In fact, they found that it took about as much time
to generate the full game tree as it did to actually solve it.  Unfortunately,
this is feasible only for small games.  In larger games where theoretically
optimal strategies cannot be computed efficiently, computer systems tend to
utilize the concept of information sets or belief states in some form or
another.  A common strategy is to estimate the value of a move by averaging the
estimated value of the associated descendant for each node in the current
information set (or a random sample from the set).  While this sort of
``averaging over clairvoyance'' {\em can} lead to very poor decisions, it is
nevertheless a reasonable strategy in some games, such as
Scrabble~\cite{sheppard02world}.

@@

\Section{Conclusion and Future Work}
We have shown that information set generation for Kriegspiel can be efficiently
parallelized.  The opportunities for parallelism increase as the problem size
increases.  Our algorithm has performed well on problems with information sets
with as many as 65 million nodes---much larger than any other treatment of the
problem that we are aware of.  Because the algorithm can be efficiently
parallelized, it would be reasonable to explore using it in a full Kriegspiel
player.  It remains to be seen whether these scalability properties hold over a
wide variety of playing styles and problem instances.  We suspect that fruitful
research remains with respect to alternative algorithms that implement variable
ordering heuristics.

>>>>

The paper is organized as follows.  In Section~\ref{Kriegspiel}, we review
necessary background material, including the rules of Kriegspiel. Next, in
Section~\ref{info} we describe the nature of the information set generation
problem and the algorithm used.  In Section~\ref{ParSSSE}, we describe our
parallel implementation, including the {\sc Charm++} Parallel State Space
Search Engine (ParSSSE) platform.  We then provide our experimental results in
Section~\ref{Results}.  Related work is discussed in Section~\ref{Related}.  We
conclude with a discussion of future work.

\section{Kriegspiel}\label{Kriegspiel}
Information Set Generation is the process of identifying the nodes in an
imperfect information game tree that are consistent with a player's
observations~\cite{richards12information}. Here, we present our techniques and
results in the context of a particular game: Kriegspiel~\cite{li94chess}.
However, the techniques and framework presented in this paper are general and
can be applied to ISG in other imperfect information games.

Kriegspiel, or partially observable chess,  was invented by Henry Michael
Temple in 1899.  Historically, the game has required three parties: two
competing players and one referee.  Today, players often compete over a
computer network, with the role of the referee being automated.  One player
sees and controls only the white pieces; the other, black.  The referee tracks
both sets of pieces.  Instead of openly declaring their moves, players secretly
communicate move requests to the referee.  The referee, who knows the location
of all pieces, assesses the legality of the move according to the standard
rules of chess.

@@

\section{Information Set Generation}
\label{info}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figure~\ref{filteredlisting},~\ref{fig:figure1} and \ref{fig:figure2} show a full listing and the associated pictorial progression of an actual
kriegspiel game~\cite{li94chess}.  We have adopted the notation used by Wolfe for Berkeley Kriegspiel.  Odd (resp. even) moves are for white (resp. black).  The actual move is shown
first followed by a list of referee announcements.  Captures announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn tries announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.  For example at move 9, (Figure~\ref{filteredlisting}),
black attempts to block a potential threat along the diagonal by moving a pawn to f6, then attempts to escape or capture
a threat at f7.  When both of these attempts fail, black can infer that his king is being threatened by a protected
bishop or queen at f7.  The attempt to move to e7 is black's way of finding out whether the threatening piece is a queen
or bishop.  When this move fails, black knows it is white's queen and ultimately retreats to d8.  Note that the failed
moves by white on turns 3 and 4 are a result of being blocked by the pawn at e2.

@@

\section{Parallel Implementation using Charm++ and ParSSSE}\label{ParSSSE}
We built our search engine on top of the {\sc Charm++} run-time
system for parallel processing~\cite{kale93charm,kale09charm}.  {\sc Charm++}
is an extension to the C++ programming language that provides
machine-independent infrastructure for parallel computation.  The language and
its accompanying runtime system have been ported to many shared-memory and
distributed-memory platforms.

@@

\section{Performance Results}\label{Results}
In order to understand the effect of various parameters such as grainsize and
load balancing strategy, we first evaluated the scalability of our
implementation using a randomly selected problem instance. Next, we show that
our implementation performs equally well for instances from actual games.

\subsection{Analysis using Random Instance} 
The performance results reported in this section were obtained by running our
program on Blueprint with up to 1,024 processors. This system was composed of
120 POWER5+ nodes, each having 16 cores and 64GB of memory. The system was
decomissioned in June 2011 and is no longer in operation.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}
\vspace{-0.1in}

@@

\label{speedups}
\vspace{-0.3in}
\end{table}
In order to understand the effect of various parameters such as grain size and load balancing, we first evaluated the scalability of our implementation using a randomly selected problem instance. Next, we show that our implementation performs equally well for instances from actual games.
\subsection{Analysis using Random Instance}
The performance results reported in
this section were obtained by running our program on Blueprint with up to 1024 processors. This system was composed of 120 POWER5+ nodes, each having 16 cores and 64GB of memory. The system was decomissioned in June 2011 and is no longer in operation.

Table~\ref{speedups} shows the performance results for our algorithm on a randomly selected problem instance in which
the size of the information set was about 4.5 million.  This instance is not necessarily representative of all games or
states of the game, but it was not cherry-picked in any way. The table shows that the problem is indeed highly
parallizable.  The speedups are nearly linear for up  to 32 processors, and even on 1024 processors, we get a good
speedup.  In this particular instance, the sequential algorithm takes almost 25 minutes to run.  But when parallelized
on 1024 processors---perfectly reasonable resource usage on modern hardware---the search takes less than three seconds.
This is a speedup of 576 times!  It might not be reasonable to use the algorithm at all if it takes 25 or more to run
for some turns.  But if the cost is only a few seconds, it becomes entirely feasible.

\subsubsection{Effect of Load Balancing}

In our next round of experiments, we set out to compare load-balancing
strategies for the problem.  We used the random, work stealing, and
neighborhood balancing strategies that are available in the general purpose
search engine.  These results are shown in Figure~\ref{3schemes}.  Our results
show that there was actually little variation in performance across the
different strategies.  Across all of our experiments, the work stealing
approach appeared to be consistently---if only marginally---better than the
others.  Our hypothesis here is that the nodes are rarely starved for work.  In
other words, perhaps the very nature of the problem means that the load is
naturally well-balanced and that the load-balancer rarely needs to do a
significant amount of redistribution.  This is probably also related to the
fact that we are doing an all-solutions search. 
\begin{figure}[ht]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}
\vspace{-0.1in}

@@

\caption{Effect of load balancing strategy}
\label{3schemes}
\vspace{-0.2in}
\end{figure}

\begin{figure*}[!ht]
\centering
\subfigure[Depth 8]{
\includegraphics[width=2.1in]{plots/dep8.pdf}
\label{dep8}
} 
\subfigure[Depth 10]{
\includegraphics[width=2.1in]{plots/dep10.pdf}
\label{dep10}
} 
\subfigure[Depth 12]{
\includegraphics[width=2.1in]{plots/dep12.pdf}
\label{dep12}
} 
\vspace{-0.1in}
\caption{Grainsize Control for different problem sizes}
%\vspace{-0.1in}
\label{fig:dep}
\vspace{-0.1in}
\end{figure*}

\subsubsection{Effect of Problem Size}

@@

\subsection{Real Game Instance}
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem4.pdf} %
%
%\caption{Kriegspiel position from Virgil-David 1969 with information set size
%764,209.}
%
%\label{prob4} 
%\vspace{-0.2in}
%\end{figure}
%
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem5.pdf} %
%
%\caption{Kriegspiel position from Game 6 of Kbot-Darkboard at the 2006 Computer
%Chess championship.  Information set size is 5,099,257.}
%
%\label{prob5} 
%\vspace{-0.2in}
%\end{figure}

\begin{figure}[ht]
 \centering \includegraphics[scale=1]{plots/new.pdf} %
\caption{Kriegspiel position from 2 games, position 10}
\label{prob4} 
\vspace{-0.2in}
\end{figure}

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 1844$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

\begin{table}[th]
\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & 617$|$1 & {\bf 612$|$1} & 650$|$0.95 & 3312$|$0.19 \\
4 & 213$|$2.89 & 170$|$3.63 & {\bf 158$|$3.89} & 367$|$1.68 \\
16 & 83$|$7.41 & 49$|$12.4 & {\bf 40$|$15.1} & 64$|$9.57 \\
64 & 54$|$11.3 & 19$|$32.3 & {\bf 11$|$52.9} & 14.9$|$41.4 \\
256 & 54$|$11.3 & 9.8$|$62.9 & 3.5$|$172 & {\bf 3.5$|$174} \\
1024 & 58$|$10.6 & 5.5$|$111 & 1.3$|$464 & {\bf 1.2$|$536} \\
\end{tabular}

@@

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrc}
{\bf Authors} & {\bf Game} & {\bf Year} & {\bf Best Speedup (cores)} \\
Ferguson \& Korf & Othello & 1988 & 12(32) \\
Felton \& Otto & Chess & 1988 & 100(256) \\
Richards \etal & Kriegspiel & 2012 & 178(256) \\
Richards \etal & Kriegspiel & 2012 & {\bf 536(1024)} \\
\end{tabular}
\caption{Best Speedups reported for game tree applications.}
\vspace{-0.2in}
\label{bestspeedups}
\end{table}

@@

\section{Related Work}
\label{Related}
An overview of the theory of imperfect information games and other basic topics
in game theory can be found in~\cite{kuhn03lectures} and~\cite{kuhn97classics}.
A technical explanation of the concept of information sets can be found
in~\cite{gilpin07algorithms}.  Koller \etal, showed that a Nash equilibrium can
be found in time that is polynomial in the number of nodes in the game
tree~\cite{koller94fast}.  In fact, they found that it took about as much time
to generate the full game tree as it did to actually solve it.  Unfortunately,
this is feasible only for small games.  In larger games where theoretically
optimal strategies cannot be computed efficiently, computer systems tend to
utilize the concept of information sets or belief states in some form or
another.  A common strategy is to estimate the value of a move by averaging the
estimated value of the associated descendant for each node in the current
information set (or a random sample from the set).  While this sort of
``averaging over clairvoyance'' {\em can} lead to very poor decisions, it is
nevertheless a reasonable strategy in some games, such as
Scrabble~\cite{sheppard02world}.

@@

\section{Conclusion and Future Work}
We have shown that information set generation for Kriegspiel can be efficiently
parallelized.  The opportunities for parallelism increase as the problem size
increases.  Our algorithm has performed well on problems with information sets
with as many as 65 million nodes---much larger than any other treatment of the
problem that we are aware of.  Because the algorithm can be efficiently
parallelized, it would be reasonable to explore using it in a full Kriegspiel
player.  It remains to be seen whether these scalability properties hold over a
wide variety of playing styles and problem instances.  We suspect that fruitful
research remains with respect to alternative algorithms that implement variable
ordering heuristics.
